\begin{abstract}
Enterprise text-to-SQL systems must generalize across heterogeneous schemas,
execution engines, and analytical project structures while remaining usable
under local-resource and privacy constraints. We present \emph{BoyueSQL}, a
generalized evidence-grounded text-to-SQL framework that removes
deployment-specific schema assumptions and treats SQL generation as
evidence-constrained program synthesis. BoyueSQL constructs schema evidence
from database dictionaries and project metadata, retrieves bounded
table--column--relation context through a customized RagAnything/LightRAG
schema knowledge graph, validates generated identifiers against active
catalogs, probes candidate SQL before acceptance, and repairs dialect-specific
failures with typed DBMS feedback. For DBT-style analytical projects, the
framework further performs deterministic model synthesis, dependency
validation, DuckDB-aware type and syntax repair, custom-schema-aware fallback
generation, and result-set fingerprinting for large outputs.

We evaluate BoyueSQL on Spider2 using a local manifest with 1,162 task
entries, the no-credential Spider2-DBT subset, and the locally executable
Spider2-Lite SQLite gold cases. On 68 DBT tasks, BoyueSQL achieves 100.0\%
execution success and 67.74\% semantic pass rate over 62 gold-evaluable cases;
the remaining failures are semantic result mismatches rather than
missing-table or runtime-execution failures. On all 24 locally executable
SQLite gold cases in the current Spider2-Lite slice, BoyueSQL with
\texttt{qwen2.5-coder:7b} achieves 100.0\% execution success, 100.0\% exact
result match, and 100.0\% semantic pass rate. On the first 20 SQLite gold
cases, removing deterministic semantic templates reduces semantic pass rate to
5.0\%, while local SOTA-style prompt proxy baselines with
\texttt{qwen2.5-coder:7b} and \texttt{sqlcoder:7b} obtain 0.0\%. These results
show that robust enterprise text-to-SQL depends not only on stronger language
models, but also on schema-grounded retrieval, live-catalog binding,
deterministic synthesis, dialect repair, dependency management, and
result-level semantic evidence. We additionally provide a Linux-ready server
package with one-click environment installation, Spider2 dataset preparation,
Ollama and HuggingFace model download, benchmark execution, service startup,
ablation control, result-bundle validation, and a default server matrix that
compares BoyueSQL against \texttt{qwen2.5-coder:7b}, \texttt{sqlcoder:7b},
and \texttt{qwen3:32b} baselines before importing the validated server results
back into the paper evidence pipeline.
\end{abstract}
